---
title: "Practical Machine Learning - Final Project Report"
author: "Chitresh Pandey"
output: html_document
---
The Project Involves analyzing the Data collected in a controlled experient on 6 subjects. The activity involved here is "Unilateral Dumbbell Biceps Curl" . The data was collected for 5 different methods for the activity denoted as ( A , B , C , D ,E) . Out of 5, only A is the correct method . 

A - Exactly as per specification

B - Throwing the elbows to the front

C - Lifting the dumbbell only halfway

D - Lowering the dumbbell only halfway

E - Throwing the hips to the front


Now, here we can pose 2 questions

Q1 . Can we classify that a person is doing the "Unilateral Dumbbell Biceps Curl"  correctly or incorrectly through  a given dataset of specific measurements ? 

Q2 . If the person is classified as doing the "Unilateral Dumbbell Biceps Curl" incorrectly , can we further classify him/her into B,C,D or E ? 


The data for analysis is provided for by 

Ugulino, W.; Cardador, D.; Vega, K.; Velloso, E.; Milidiu, R.; Fuks, H. Wearable Computing: Accelerometers' Data Classification of Body Postures and Movements. Proceedings of 21st Brazilian Symposium on Artificial Intelligence. Advances in Artificial Intelligence - SBIA 2012. In: Lecture Notes in Computer Science. , pp. 52-61. Curitiba, PR: Springer Berlin / Heidelberg, 2012. ISBN 978-3-642-34458-9. DOI: 10.1007/978-3-642-34459-6_6. 


Loading the necessary libraries and the analysis data . Once the Experiment data is loaded dividing the data into Training and Testing samples . 

```{r,echo=FALSE}
setwd("C:/CHITRESH - DUMPS/COURSEERA - DATA SCIENTIST TOOLBOX/MACHINE LEARNING")
```


```{r,echo=FALSE,message=FALSE}
# Loading necessary libraries
library(ggplot2)
library(caret)
library(randomForest)
library(mboost)
library(splines)
```

Loading the Experiment data and slicing it into Testing and Training Sets
```{r}
 
ExpData <- read.csv("./pml-training.csv") 
set.seed(4868086)
inTrain <- createDataPartition( y = ExpData$classe , p = 0.7 , list = FALSE)
Training <- ExpData[inTrain , ]
Testing <- ExpData[-inTrain , ]
rm("inTrain")

```

Now creating a Transform function to manage the tranformations that the Training and Testing data may have to undergo. 

```{r,echo=FALSE }

Transform <- function(DataSet)
        {
                DataSet[is.na(DataSet)] <- 0
                s_data <- subset(DataSet , 
                                 select = -c(X,user_name,raw_timestamp_part_1,raw_timestamp_part_2,cvtd_timestamp))
                class <- s_data$classe
                s_data <- subset(s_data , select = -classe)
                data <- cbind(class , s_data)
                return(data)
        }
```

Once we have the Transformed Training and Testing data , identifying the near zero covariates for both Training data and removing them from both Training and Testing dataset.


```{r}
Training <- Transform(Training)
Testing  <- Transform(Testing)

# Identifying the Near Zero CoVariates in Training set and removing them from both Training and Testing data

NZV <- nearZeroVar(Training , saveMetrics=TRUE)
Training <- Training[ ,!NZV$nzv]
Testing  <- Testing[  ,!NZV$nzv]
```

Now , preparing my model on the training data  . 

```{r,echo=FALSE,message=FALSE}

ModelFit <- randomForest(class ~ . , 
             data = Training[floor(runif(1000 , min = 1 , max = dim(Training)[1])) , ])
```


```{r}
ModelFit1 <- update(ModelFit , data = Training) # Random Forest 

```


```{r, echo=FALSE,message=FALSE}
ModelFit <-   train(class ~ . , 
                    method  = "gbm",
                    data = Training[floor(runif(1000 , min = 1 , max = dim(Training)[1])) ,],
                    verbose = FALSE)
```

```{r}
ModelFit2 <- update(ModelFit ,method  = "gbm",data = Training,verbose = FALSE) # Boosting

```


Following are the statistics for both the Model Fits . 

```{r}
print(ModelFit1) # Random Forest 
print(ModelFit2) # Boosting 

```

Now predicting the output for the Testing data sets from the Models 

```{r}

ResultSet1 <- predict(ModelFit1 , Testing)
ResultSet2 <- predict(ModelFit2 , Testing)

```

Evaluating the Confusion Matrix for the two Models to compare the estimated Out of Sample error rates 

```{r}
confusionMatrix(Testing$class , ResultSet1)
confusionMatrix(Testing$class , ResultSet2)
```


On comapring the confusion Matrix for both the Model Fit , we observe that the Random Forest provides a better prediction estimates and lower out of sample error rates as compared to Boosting . Hence , implementing the Random Forest Model to our validation dataset ( ModelFit1 )

Now , as we have finalized our Model , importing the data for Validation 

```{r}
Validation <- read.csv("./pml-testing.csv")
ValResult <- predict(ModelFit1 , Validation)
```

Submitted the valdiation results in the necessary format.


